Fractional Types

### 2016-12-18 typespp.tex
synchronized pointed types

@@

Each combinator $\rho$ has an inverse.
\caption{$\Pi^/$ fractional combinators with pointed types (dependent version)}
\label{pifracd:comb}
\end{figure}

%%%%%
\subsection{Pointed Types}

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to modify
the operational semantics to use a mechanism to express a computational effect
which enables spatially separated parts of the program to communicate in a way
that is reminiscent of entanglement in quantum mechanics. Possibilities to
realize this communication are global reference cells, backtracking, logical
variables with unification. We choose a more abstract and enlightening approach
that encodes the required dependency in dataflow constraints using dependent
types.

\begin{definition}[Dependent $\Pi^/$ with pointed types]
\label{def:pifracd}
The syntax of dependent $\Pi^/$ with pointed types differs from
Def.~\ref{def:pifrac} as follows:
\[\begin{array}{lrcl}
(\textrm{Pointed Fractional Types}) & \mathbb{T_\bullet} &::=& [\mathbb{T},v] \\
 (\textrm{Pointed /-combinators}) & \rho_\bullet &:&
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2] ~::=
  [\textit{see Fig.~\ref{pifracd:comb}}]
\end{array}\]
\end{definition}

>>>>

\begin{figure}[t]
\[\begin{array}{rrcll}
\unitetl/ :&  \order{\idiso} \circledast \mathbb{T} & \leftrightspoon & \mathbb{T} &: \unititl/ \\
\unitetr/ :&  \mathbb{T} \circledast \order{\idiso} & \leftrightspoon & \mathbb{T} &: \unititr/ \\
\swapt/ :&  \mathbb{T}_1 \circledast \mathbb{T}_2 & \leftrightspoon & \mathbb{T}_2 \circledast \mathbb{T}_1 &: \swapt/ \\
\assoclt/ :&  \mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3) & \leftrightspoon & (\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3
  &: \assocrt/ \\
\\
\eta-_k :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  [\iorder{c},1/c^k] \circledast [\order{c},c^k] &: \epsilon-_k \\
\eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
\end{array}\]

@@

Each combinator $\rho$ has an inverse.
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

%%%%%
\subsection{Dependent Pointed Types}

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to modify
the operational semantics to use a mechanism to express a computational effect
which enables spatially separated parts of the program to communicate in a way
that is reminiscent of entanglement in quantum mechanics. Possibilities to
realize this communication are global reference cells, backtracking, logical
variables with unification. We choose a more abstract approach that encodes the
required dependency in dataflow constraints using dependent types.

\begin{definition}[Dependent $\Pi^/$]
\label{def:pifracd}
The syntax of dependent $\Pi^/$ differs from Def.~\ref{def:pifrac} as follows:
\[\begin{array}{lrcl}
(\textrm{Fractional Types}) & \mathbb{T_\bullet} &::=&
  [\order{c},c^k] \alt
  [\iorder{c},1/c^k] \alt
  [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
  \exists k.\mathbb{T_\bullet} \\
 (\textrm{/-combinators}) & \rho_\bullet &:&
  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
  [\textit{see Fig.~\ref{pifracd:comb}}]
\end{array}\]
\end{definition}

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We then allow indices of values $k$ to be range over a collection
of types. In particular, the type
$\exists k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.
